Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data2.Rmd
update table 1 and 2 tables in appendix
Please just refer to changes in tables, other changes i did was to solve running problems

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, 
      | Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

@@

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

| **Concept**    | **Details**                                                                |
|----------------|----------------------------------------------------------------------------|
| MCAR           | Missing Completely At Random, where the probability to be missing is equal |
|                | across all data entries                                                    |
| MAR            | Missing At Random, where the probability to be missing depends on observed |
|                | information                                                                |
| MNAR           | Missing Not At Random (MNAR), where the probability to be missing          |
|                | depends on unrecorded information, making the missingness non-ignorable    |
|                | [@rubi76; @meng94].                                                        |
|                |                              |

@@

<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->
<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->

@@

<!-- TODO: explicit statement about not going into workings of the methods. Galimer 2l methods. -->

## Setup
<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
devtools::install_github("hanneoberman/miceheckman")
```

<!-- [TODO: Add environment info, seed and version number(s) somewhere.]  -->

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

<!-- [TODO: explain case study] -->

In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice. We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$). The following variables are of primary interest:

@@

<!-- TODO: choose sex or gender -->

<!-- $$ -->
<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->
<!-- $$ -->
```{r}
mod <- popular ~ 1 + sex + (1 | school)
```

@@

@@

First we plot the pattern of missing data within categories of the relevant variables. Plot the missing data pattern:

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data", fig.margin=TRUE}
plot_pattern(popmis)
```

@@

<!-- This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")). -->
<!--  [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?] -->

@@

<!-- ```{r pop-cca, echo=FALSE} -->
<!-- est_cca <- lme4::lmer(mod, popmis) %>% broom.mixed::tidy() -->
<!-- # as.matrix(est_cca$estimate) - as.matrix(est_true) -->
<!-- est_cca$estimate - est_true$estimate -->
<!-- ``` -->

@@

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

@@

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

<!-- TODO: explain what this function does -->

@@

<!-- [TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.] -->

@@

```{r pop_predictor, message=FALSE, warning=FALSE, cache=TRUE}
# adjust the predictor matrix
pred["popular", "school"] <- 1 
plot_pred(pred)

@@

Analyze the imputations:
```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

@@

```{r pop_multilevel, message=FALSE, warning=FALSE, cache=TRUE}
# adjust the predictor matrix
pred["popular", "school"] <- -2 
plot_pred(pred)

@@

<!-- [TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.] -->

We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc]. 
<!-- [TODO: add more info about the complete data.]  -->
The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, 
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

<!-- [TODO: make ct levels one var? also shows that you don't always need random effects everywhere?] -->

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

@@

Load the complete and incomplete data into the R workspace:
```{r data, eval = FALSE}
data("impact", package = "metamisc")      # complete data
dat <- read.table("link/to/the/data.txt") # incomplete data
```

@@

fits %>% left_join(n, by = "name") %>% 
  ggplot(aes(y = name, x = estimate, color = term, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```
<!-- We will use the following estimates as comparative truth in this tutorial: -->
```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

<!-- [TODO: show how much variance there is after different methods] -->

<!-- [TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.] -->

<!-- [TODO: add descriptive statistics of the complete and incomplete data.] -->

@@

To explore the missingness, it is wise to look at the missing data pattern. The ten most frequent missingness patterns are shown:
```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
```

This shows that we need to impute `ct` and `pupil`.

To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. See also all of the TODOs that need to be worked out. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

<!-- Missing data pattern with clustering -->
<!-- TODO: simplify argument in md pattern plot and/or network plot for md patterns (summarized per variable?) -->
<!-- TODO: test if there's a random distribution of missingness in the md pattern? -->
<!-- TODO: check spss missing data pie chart? -->
<!-- TODO: similarity of patterns over clusters -->
<!-- TODO: gradient (alpha) to show in how many of the clusters the row in the pattern occurs -->
<!-- TODO: 3 columns could be 3 boxplots, over these clusters we have a distribution of these statistics across clusters.  -->
<!-- TODO: look at vim -->
<!-- TODO: check if shepley plot is an option -->

Many datasets include individuals from multiple settings, geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). In general, individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. [TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation. VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

<!-- Maybe besides CC and imputation name also likelihood based methods -->
Like any other dataset, clustered datasets are prone to missing data. Several strategies can be used to handle missing data, including complete case analysis and imputation. We focus on the latter approach and discuss statistical methods for replacing the missing data with one or more plausible values. Afterwards, the completed data can be analyzed as if they were completely observed. In contrast to single imputation (where missing data are only replaced once), multiple imputation allows to preserve uncertainty due to missingness and is therefore recommended (c.f. Rubin 1976).

When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. [TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing. [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"]

@@

| **Concept**          | **Details**                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      |
|-----------------|-------------------------------------------------------|
| Sample unit          | Units of the population from which measurements are taken in a sample.                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
| Hierarchical levels  | Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.                                                                                                                                                                                                                                                                                                                                                                                    |
| Fixed effect         | Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.                                                                                                                                                                                                                                                                                                                                                                             |
| Random effect        | Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.                                                                                                                                                                                                                                                                     |
| Mixed effect         | Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters [see e.g. @gelm06, @hox17 and @jong21].  |
| ICC                  | The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.                                                                                                                                                                                                                                      |
| Stratified intercept |                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                  |

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```

<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

Imputation of missing data requires consideration of the mechanism behind the missingness. Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}). For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15. We here consider the general case that data are MAR, and expand on certain MNAR situations.

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:miss}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
& across all data entries \\
MAR     & Missing At Random, where the probability to be missing depends on observed \\
& information \\
MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
& depends on unrecorded information, making the missingness non-ignorable \\
& [@rubi76; @meng94]. \\
& [TODO: add congeniality, but maybe in-text?] \\
\hline
\end{tabular}
\end{table}

The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.  For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.

[TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.]

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial. [TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g. @audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].

- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
- `hiv` from the \pkg{GJRM} package [simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions with data that are MNAR, @GJRM].

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).

@@

| **Concept** | **Details**                                                                                                                                                   |
|----------------|--------------------------------------------------------|
| MCAR        | Missing Completely At Random, where the probability to be missing is equal across all data entries                                                            |
| MAR         | Missing At Random, where the probability to be missing depends on observed information                                                                        |
| MNAR        | Missing Not At Random (MNAR), where the probability to be missing depends on unrecorded information, making the missingness non-ignorable [@rubi76; @meng94]. |

@@

<!-- [TODO: clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implementation ("thou shall not simply run `mice()` on any incomplete dataset").] -->

<!-- [TODO: Add that the more the random effects are of interest, the more you need multilevel imputation models.] [TODO: Add an overview of all possible predictor matrix values in manuscript or `ggmice` legend.] -->

@@

TODO: explicit statement about not going into workings of the methods. Galimer 2l methods.

## Setup

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:

<!-- * TODO: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB? -->

Install non-CRAN packages if necessary:

```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
devtools::install_github("hanneoberman/miceheckman")
```

Set up the R environment and load the necessary packages:

```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(miceadds)     # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
library(miceheckman)       # for imputation cf. heckman models
library(metamisc)     # for case study data

```

<!-- library(broom.mixed)  # for multilevel estimates -->

@@

[TODO: explain case study]

In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice. We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$). The following variables are of primary interest:

@@

<!-- $$ -->

<!-- \texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)} -->

<!-- $$ -->

```{r}
mod <- popular ~ 1 + sex + (1 | school)
```

@@

The true effect is:

```{r echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(popcomp$school),
    ~ {
      lm(
        mod,
        data = filter(popcomp, school == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(school = as.character(.x), .)
  ) %>% filter(term %in% "sex")

n <- popcomp %>% group_by(school) %>% 
  tally() %>% 
  mutate(school = as.character(school))

fits %>% left_join(n, by = "school") %>% 
  ggplot(aes(y = school, x = estimate, size = n)) +
  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
```

@@

Plot the missing data pattern:

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis)
```

@@

<!-- This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")). -->

<!--  [TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?] -->

@@

<!-- ```{r pop-cca, echo=FALSE} -->

<!-- est_cca <- lme4::lmer(mod, popmis) %>% broom.mixed::tidy() -->

<!-- # as.matrix(est_cca$estimate) - as.matrix(est_true) -->

<!-- est_cca$estimate - est_true$estimate -->

<!-- ``` -->

@@

Impute the data, ignoring the cluster structure:

```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
imp <- mice(popmis, pred = pred, print = FALSE)
```

@@

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex  + (1 | school))) 
```

@@

[TODO: Add that it doesn't work with systematic missingness (only with sporadic). There's some pros and cons, and it may not even differ much if the number of clusters is low.]

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred["popular", "school"] <- 1 
plot_pred(pred)

@@

Analyze the imputations:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
fit <- with(imp, 
            lmer(popular ~ 1 + sex + (1 | school))) 
```

@@

```{r pop_multilevel, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred["popular", "school"] <- -2 
plot_pred(pred)

@@

[TODO: check if there is systematic missingness in this dataset, if not make Marshall Computerized Tomography classification (ct) systematically missing.]

We illustrate how to impute incomplete multilevel data by means of a case study: `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc]. [TODO: add more info about the complete data.] The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model:

```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

@@

Load the complete and incomplete data into the R workspace:

```{r data, eval = FALSE}
data("impact", package = "metamisc")      # complete data
dat <- read.table("link/to/the/data.txt") # incomplete data
```

@@

fits %>% left_join(n, by = "name") %>% 
  ggplot(aes(y = name, x = estimate, color = term, size = n)) +
  geom_vline(xintercept = 1, size = 1.5, color = "grey95") +
  geom_linerange(aes(y = name, xmin = conf.low, xmax = conf.high, size = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "Adjusted odds ratio for ct", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
# TODO: add heterogeneity in distribution of ct -> prevalence in ct across clusters with CI
# TODO: missingness prop inside clusters
# TODO: make row height proportional to pattern frequency
```

<!-- We will use the following estimates as comparative truth in this tutorial: -->

```{r truth}
# fit <- glmer(mod, family = "binomial", data = impact) # fit the model
# tidy(fit, conf.int = TRUE, exponentiate = TRUE)       # print estimates
```

[TODO: show how much variance there is after different methods]

[TODO: add ICC before/after imputation and interpret: This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.]

[TODO: add descriptive statistics of the complete and incomplete data.]

@@

To explore the missingness, it is wise to look at the missing data
pattern. The ten most frequent missingness patterns are shown:

To explore the missingness, it is wise to look at the missing data pattern:

```{r pattern, fig.height=7.1}
plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
```

This shows... [TODO: fill in that we need to impute `ct` and `pupil`.]

To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we... [TODO: fill in that we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).]
